Discussion Draft

### 2020-04-15 doc/Discussion.rmd
cleaning up sentences

Under the spindle based model, faster male evolution of the genome wide recombination rate, is primarily driven by a more efficient spindle assembly checkpoint (SAC) acting on chromosome structures at metaphase I alignment. <We refer the reader to [@laneKauppi2019] for more details on the SAC pathway.> The SAC prevents anueploidy by ensuring all bivalents are correctly attached to the microtubulue spindle before starting the irreversible metaphase-to-anaphase transition (cite).  A stable chromosome structure, through bi-orientation, requires, i) kinetochores attached to opposite poles of the cells and ii) at least one crossover to create tension across a proportion of sister cohesion (cite). Unstable bivalents on the meiosis I spindle, cause the SAC to delay anaphase and allow time for the bivalent to stabilize on the microtubule spindle. If they fail to attach, the SAC triggers apoptosis. Genetic variants which increase the number of unstable bivalents (such as those affecting recombination rate or cohesion), will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle. Multiple lines of evidence indicate that the effectiveness of SAC in spermatogenesis is greater compared to oogenesis (LaneKauppi, Subrahman). 
These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle and cell volume (cite). This model nominates conserved differences in in the context of meiosis between males and females as the primary source of distinct evolutionary trajectories in genome wide recombination rates.

Regardless of if these hypotheses turn out to be true, our results extend previous reports [@dumont2011] of rapid male-specific evolution in multiple subspecies and strains of house mouse. Whether these instances of evolution are due to separate independent events or segregating genetic variation within the population requires further study. The faster male evolution in genome-wide recombination rates suggests that even for species with female biased heterochiasmy, selecting for higher male specific genome wide recombination rates may be the most effective way to increase the sex-averaged recombination rate for a genetic background. This insight could be important for animal breeding programs [@battagin2016].

@@

Our results of greater between cell variation in females are supported by empirical results from human and mouse [@lynn2002, @gruhn2013].

While the modifier models parameterize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant.

In the spindle based selection model, the less efficient SAC in oocytes leads to relaxed selection on the metaphase I chromosome structure. This will result in increased variance in chromosome structure and crossover number per cell, not just across strains, but across cells within individuals. Oocytes at metaphase I, have an acentriolar spindle comprised of  diffuse network of microtubules with multiple microtubule organizing centers (Schuh Ellenburg 2007). This is another way the spindle based selection model is linked to conserved sex differences in gametogenesis (cite).

@@

0. Reguardless of whether these models and hypothese are true if the pattern of females having more across cell variation in gwRR

1. It has a consequence of fitting with distinct evolutionary streategies between the sexes of females diversifying the genetics their offspring and males limiting the shuffling of successful combinations (Trivers).

@@

#as first outlined in petkov et al and restated in Haenel et al, -- chromatin organization

# source could be DSB formation along chromosome - and /or biases in DSB repair

#CO suppresion, elevated DSB (most repaired as NCO) -- elevated DsB (Brick)

Our results on the broad scale pattern of crossover placement are supported by the literature. As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a highly conserved trait. The two locus modifier and SACE model predicts sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks under linkage in males respectively.

The two mechanical models both predict sexual dimorphism in the recombination landscape for distinct reasons. The COM model, predicts the sex differences in the recombination landscape via interference is due to a combination of the length of the axis and differing strengths of chromosome movements during the pairing process [@hulten2011_COM, @rubin2020mixing].

@@

If these hypotheses concerning evolution of these patterns **are not true**  there's still predictions that can be made given the consequences of these patterns. The consequences of this difference in the typical landscape as reviewed in [@sardell_sex_2020]. For the sex averaged landscape along chromosomes, any diviation from uniform placement is generally due to male specific patterns.

There's also the prediction that females have active CO supression near telomeres (variation) -- geiven the evidence in (female mice) that show a higher elevation of DSBs at the telomere (halldorsson?).

@@

As outlined in @haenel2018, sexual dimorphism in chromatin organization translates to interference strength in base pairs. Since more DNA is sequestered into the radial loops per unit of the central axis, in the chromosome organization with longer DNA loops, as seen in males for mice [@petkov2007, @lynn2002], and humans [@gruhn2013], interference strength measured in base pairs will be stronger compared to female linkage maps. The signal for interference is mediated though the chromosome axis [@zicklerKleckner2015], interference strength measured in micrometers of is conserved between sexes [@petkov2007]. The wide spread pattern of stronger interference strength in Mb (and cm/Mb) for male specfic linkage maps is indirect evidence that sexual dimorphism in chromatin organization however this has yet to be confirmed in cytological data from both sexes in species beyond mice and humans.

@@

None of these models may be appropriate due to X Y and  Z caveats. The consequence of this pattern of greater interference strength is ---
 consequence of rapid evolution of normalized interference strength, larger linkage blocks.

@@

importance of evolutionary distinct trajectories, measure both sexes -- do evolutionary comparisons within sex.

wild samples, short times scales.

single cell measures

chromatid interference

integrate physical models -- centromere and kinetochore

Diversity of systems for overcoming limitations

1. identifying indiviual (autosomes) Chromosomes across cells: FISH (diffecult) or karyotypes with identifyable chromosomes (either into groups or largest)(birds, zfish?, peromyscus, Rbt mice, bovids?, humans, ), translocations (robertsonian translocations),

2. non-model orgs can be used to study meiosis -- look at the weird ones (protezoa)

@@

<a partial pattern seen in Arabidopsis thalinia (Drouaud et al 2007) however there is no SC based measure of interference and low number of double crossovers in the female specific linkage map. It also could be that the greater error in regulation of crossovers may have driven the longer female axis to ensure increase the probability of successful maturation of DSB into crossover chiasmata>

>>>>

Under the spindle based model, faster male evolution of the genome wide recombination rate, is primarily driven by a more efficient spindle assembly checkpoint (SAC) acting on chromosome structures at metaphase I alignment. <We refer the reader to [@laneKauppi2019] for more details on the SAC pathway.> The SAC prevents anueploidy by ensuring all bivalents are correctly attached to the microtubulue spindle before starting the irreversible metaphase-to-anaphase transition (cite).  A stable chromosome structure, through bi-orientation, requires, i) kinetochores attached to opposite poles of the cells and ii) at least one crossover to create tension across a proportion of sister cohesion (cite). Unstable bivalents on the meiosis I spindle, cause the SAC to delay anaphase and allow time for the bivalent to stabilize on the microtubule spindle. If they fail to attach, the SAC triggers apoptosis. Genetic variants which increase the number of unstable bivalents (such as those affecting recombination rate or cohesion), will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle. Multiple lines of evidence indicate that the effectiveness of SAC in spermatogenesis is greater compared to oogenesis ([@laneKauppi2019], Subrahman). 
These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle and cell volume (cite). This model nominates conserved differences in in the context of meiosis between males and females as the primary source of distinct evolutionary trajectories in genome wide recombination rates.

Our results extend previous reports [@dumont2011] of rapid male-specific evolution in multiple subspecies and strains of house mouse. Whether these instances of evolution of genome-wide recombination rates are due to independent events or segregating genetic variation within the population requires further study. Regaurdless, the faster male evolution in genome-wide recombination rates suggests that even for species with female biased heterochiasmy, selecting for higher male specific genome wide recombination rates may be the most effective way to increase the sex-averaged recombination rate for a genetic background. This insight could be important for animal breeding programs [@battagin2016].

@@

Our results of greater between cell variation in females are supported by empirical results from human and mouse [@lynn2002, @gruhn2013]. While the modifier models parameterize sexual dimorphism in the strength of selection which can shape the variance in a trait these types of models make predictions  for variance between individuals but not within individuals where the genetics are constant.

In the spindle based selection model, the less efficient SAC in oocytes leads to relaxed selection on the metaphase I chromosome structure. This will result in increased variance in chromosome structure and crossover number per cell, not just across strains, but across cells within individuals. Oocytes at metaphase I, have an acentriolar spindle comprised of  diffuse network of microtubules with multiple microtubule organizing centers [@schuh2007]. This is another way the spindle based selection model is linked to conserved sex differences in gametogenesis (cite).

@@

Regardless of whether these models and hypotheses are true, our results further support the results of females having greater variation in the number of crossovers across cells. This feature of heterochiasmy supports evolutionary theories predicting distinct reproductive strategies between females (diversifying offspring) and males (maintain successful haplotypes) [@trivers2002].

@@

# source could be DSB formation along chromosome - and /or biases in DSB repair 
#CO suppresion, elevated DSB (most repaired as NCO) -- elevated DsB (Brick)
```

As reviewed in [@sardell_sex_2020], sexual dimorphism in the broad scale recombination landscape is a highly conserved trait, which our results support. The two locus modifier and SACE models predict sexual dimorphism in the broad scale recombination landscape, for diminishing the effect of drive systems and maintaining larger chromosome blocks under linkage in males respectively.

The COM model, predicts the sex differences in the recombination landscape via interference is due to a combination of the length of the axis and differing strengths of chromosome movements during the pairing process [@hulten2011_COM, @rubin2020mixing].

@@

A consequence of the sexual dimorphism in broad scale recombination landscapes is that the male recombination pattern drives divation from an uniform distribution in any sex averaged recombination landscape [@sardell_sex_2020]. Elevation at the chromosome ends is a conserved trait across animals and plants and has a consequence of higher genetic diversity near chromosome ends [@haenel2018].

@@

As outlined in @haenel2018, sexual dimorphism in chromatin organization translates to interference strength in base pairs. Since more DNA is sequestered into the radial loops per unit of the central axis, chromatin organization with longer DNA loops, as seen in males for mice [@petkov2007, @lynn2002] and humans [@gruhn2013], interference strength measured in base pairs will be stronger compared to female linkage maps. The signal for interference is mediated though the chromosome axis [@zicklerKleckner2015], interference strength measured in micrometers of is conserved between sexes [@petkov2007]. Stronger interference strength in basepairs for male specfic linkage maps is indirect evidence that sexual dimorphism in chromatin organization however this has yet to be confirmed in cytological data from both sexes in species beyond mice and humans.

@@

An expected consequence of our observed elevated genome-wide recombination rate is lower LD, which may increase the efficiency of selection and impact the dynamics of intrgresion. However given that the higher genome-wide recombination rates are paired with stronger interference these predicted pattern may have heterogenous signal along chromosomes.

@@

We make three suggestions for future steps in studying recombination variation. First consider the when comparing sex specific meiotic traits consider that evolutionary distinct trajectories due to the fundamental differences in gametogenesis. <while comparisons of traits across sex -are interesting - -- some comparisons within sex are more appropriate for some evolutionary comparisons.>

Second recognize that certain systems can be leveraged to overcome limitations inherent in some approached. For example, identifying chromosomes across cells in cytological data requires chromosome specific probes, (e.g. FISH) but chromosome specific data can be collected from organisms with diverse karyotypes (birds, peromyscus,  humans) or backgrounds with Robertsonian translocations.

Third we promote the cross-pollination of physiology based models and more abstract model such as population genetic models with testablle hypotheses (Dapper Paysuer). A goal of such merging of models could be to connect empiracal findings across scales -- for example fine scale hotspot data to chromosome wide variation in recombination patterns (Keeney).

@@
